Carbonate chemistry at Ny-Ålesund

### 2023-01-05 awipev-CO2.Rmd
played with new function cunit
revised function cunit
Takahashi monthly added

@@

```{r, include=FALSE}
# According to Takahashi et al. (2002) and McCutcheon et al. (2021)
# Select data
taka <- d_all %>% # just for this chunk
  dplyr::select(datetime, monthd, sal_fb_mix, temp_insitu_11m, pco2_fb_insi_temp) %>% 
  dplyr::mutate(year = lubridate::year(datetime),
                season = dplyr::case_when(monthd %in% c(12, 1, 2) ~ "winter",
                                          monthd %in% c(3, 4, 5) ~ "spring",
                                          monthd %in% c(6, 7, 8) ~ "summer",
                                          monthd %in% c(9, 10, 11) ~ "fall")) %>% 
  # eliminate years 2015 and 2021 which are incomplete
  dplyr::filter(datetime > as.POSIXct("2016-01-01 00:00:00") &
                  datetime < as.POSIXct("2021-01-01 00:00:00"))

# annual 
taka_year <- taka %>% 
  group_by(year) %>% 
  dplyr::mutate(pco2_thermal = 
                  mean(pco2_fb_insi_temp, na.rm = TRUE) * exp(0.0423 * (temp_insitu_11m - mean(temp_insitu_11m, na.rm = TRUE))),
                  pco2_nonthermal = pco2_fb_insi_temp * exp(0.0423 * (mean(temp_insitu_11m, na.rm = TRUE) - temp_insitu_11m))) %>% 
  dplyr::summarise(thermal = max(pco2_thermal, na.rm = TRUE) - min(pco2_thermal, na.rm = TRUE),
                   nonthermal = max(pco2_nonthermal, na.rm = TRUE) - min(pco2_nonthermal, na.rm = TRUE),
                   thermal_nonthermal = thermal / nonthermal) %>% 
  ungroup()

@@

`r knitr::kable(taka_year, digits = c(0, 1, 1, 2), caption = "Thermal vs non-thermal control of pCO2. Yearly estimate.", format = "simple")`
`r knitr::kable(taka_season, digits = c(0, 1, 1, 2), caption = "Thermal vs non-thermal control of pCO2. Seasonal estimate.", format = "simple")`
For seasonal analysis, December to February was considered winter, March
to May was considered spring, June to August was considered summer, and
September to November was considered fall.

The control of pCO2 by non-thermal (physical and biological processes)
drivers is always greater than the control from temperature on both
annual and seasonal basis. Thermal control is highest in winter and
lowest in summer.

>>>>

#################### Function cunit
# cunit(d = 65,i = "g", o = "mol")
# cunit(d = 65,i = "g", o = "mol")
cunit <- function(d = 65, i = "gC", o = "molCO2") {
    if (i == "gC" & o == "molCO2") {
      out <- d/12
      attributes(out) <- list(unit = "mol CO2")
    }
    if (i == "molC" & o == "molCO2") {
      out <- d*44/12
      attributes(out) <- list(unit = "mol CO2")
    }
  return(round(out, digits=2))
}

@@

```{r, include=FALSE}
# According to Takahashi et al. (2002) and McCutcheon et al. (2021)
# Select data
taka <- d_all %>% # just for this chunk
  dplyr::select(datetime, montht, monthd, sal_fb_mix, temp_insitu_11m, pco2_fb_insi_temp) %>% 
  dplyr::mutate(year = lubridate::year(datetime),
                season = dplyr::case_when(monthd %in% c(12, 1, 2) ~ "winter",
                                          monthd %in% c(3, 4, 5) ~ "spring",
                                          monthd %in% c(6, 7, 8) ~ "summer",
                                          monthd %in% c(9, 10, 11) ~ "fall")) #%>% 
  # # eliminate years 2015 and 2021 which are incomplete
  # dplyr::filter(datetime > as.POSIXct("2016-01-01 00:00:00") &
  #                 datetime < as.POSIXct("2021-01-01 00:00:00"))

# monthly 
taka_month <- taka %>%
  group_by(montht) %>%
  dplyr::mutate(
    pco2_thermal =
      mean(pco2_fb_insi_temp, na.rm = TRUE) * exp(0.0423 * (
        temp_insitu_11m - mean(temp_insitu_11m, na.rm = TRUE)
      )),
    pco2_nonthermal = pco2_fb_insi_temp * exp(0.0423 * (
      mean(temp_insitu_11m, na.rm = TRUE) - temp_insitu_11m
    ))
  ) %>%
  dplyr::summarise(
    thermal = max(pco2_thermal, na.rm = TRUE) - min(pco2_thermal, na.rm = TRUE),
    nonthermal = max(pco2_nonthermal, na.rm = TRUE) - min(pco2_nonthermal, na.rm = TRUE),
    thermal_nonthermal = thermal / nonthermal
  ) %>%
  ungroup()
# monthly plot
taka_month_plt <- taka_month %>% 
  ggplot() +
  geom_point(aes(x = montht, y = thermal_nonthermal), col="blue", size = 2) +
  labs(x="Month", y= "Thermal/non-thermal control of pCO2")  +
  geom_hline(yintercept = 1, linetype = "dashed") +  
  theme_bw() + Mytheme(size_labs = 12) + theme(aspect.ratio = 1) +
  coord_fixed(ratio = 1, expand = TRUE, clip = "on")
taka_month_plt
ggsave(filename = "figures/taka_month.png", plot = taka_month_plt, width = 25, height = 25, units = "cm")

# annual 
taka_year <- taka %>%
  # eliminate years 2015 and 2021 which are incomplete
  dplyr::filter(datetime > as.POSIXct("2016-01-01 00:00:00") &
                  datetime < as.POSIXct("2021-01-01 00:00:00")) %>% 
  group_by(year) %>% 
  dplyr::mutate(pco2_thermal = 
                  mean(pco2_fb_insi_temp, na.rm = TRUE) * exp(0.0423 * (temp_insitu_11m - mean(temp_insitu_11m, na.rm = TRUE))),
                  pco2_nonthermal = pco2_fb_insi_temp * exp(0.0423 * (mean(temp_insitu_11m, na.rm = TRUE) - temp_insitu_11m))) %>% 
  dplyr::summarise(thermal = max(pco2_thermal, na.rm = TRUE) - min(pco2_thermal, na.rm = TRUE),
                   nonthermal = max(pco2_nonthermal, na.rm = TRUE) - min(pco2_nonthermal, na.rm = TRUE),
                   thermal_nonthermal = thermal / nonthermal) %>% 
  ungroup()

@@

`r knitr::kable(taka_year, digits = c(0, 1, 1, 2), caption = "Thermal vs non-thermal control of pCO2. Yearly estimate.", format = "simple")`
`r knitr::kable(taka_season, digits = c(0, 1, 1, 2), caption = "Thermal vs non-thermal control of pCO2. Seasonal estimate.", format = "simple")`
`r knitr::kable(taka_month, digits = c(0, 1, 1, 2), caption = "Thermal vs non-thermal control of pCO2. Monthly estimate.", format = "simple")`
![Monthly analysis of the thermal vs non-thermal control of pCO2](figures/taka_month.png)

The relative importance of the temperature and biological effects maybe expressed in terms of the ratio,(temperature effect)/(biology effect) or difference [(temperature effect)-(biological effect)].

For annual analysis, only complete years were used. For seasonal analysis, December to February was considered winter, March to May was considered spring, June to August was considered summer, and
September to November was considered fall.

The control of pCO2 by non-thermal (physical and biological processes)
drivers is always greater than the control from temperature on both
annual and seasonal basis. Thermal control is highest in winter and
lowest in summer. This is very clear with the monthly analysis.
